---
title:  "Sleep Study on the Job"
subtitle: "An investigation into sleep qualities and occupation." 
date: today
date-format: long 
author: 
 - name: Kadin Willkins 
   email: kadinw@berkley.edu 
   affiliations: 
   - name: UC Berkeley, School of Information 
     address: 102 South Hall Drive
     city: Berkeley, CA 
     postal-code: 94720
 - name: Kenny Jeong
   email: 
   affiliations: 
   - name: UC Berkeley, School of Information
     address: 102 South Hall Drive
     city: Berkeley, CA 
     postal-code: 94720
 - name: Oleksii Lavrenin
   affiliations: 
   - name: UC Berkeley, School of Information
     address: 102 South Hall Drive
     city: Berkeley, CA 
     postal-code: 94720
format: 
  pdf: 
    documentclass: scrartcl
toc: false 
colorlinks: true
cap-locaiton: bottom
execute: 
  echo: false 
  warning: false 
  message: false 
bibliography: references.bib
citeproc: false
biblio-title: References
link-citations: true
abstract: | 
  <center>A deep dive into the correlation between sleep quality and occupation</center>
  

---
\newpage
```{r}
library(corrplot)
library(tidyverse)
library(patchwork)

library(sandwich)
library(stargazer)

library(ggplot2)
library(broom)
library(nnet)
library(dplyr)
```
```{r set plotting theme}
theme_set(theme_minimal())
sleep_data <- read.csv("Sleep_health_and_lifestyle_dataset.csv")
```
```{r}
get_robust_se <- function(vcov) {
  se_ <- sqrt(diag(vcov))
  return(se_)
}
```
## Importance and Context
In our study we aim to explore potential connections between sleep quality and occupation using a dataset containing occupations, sleep duration, stress levels, BMI, blood pressure, daily steps, gender, age, physical activity, and finally sleep quality. Our primary question is: “Is there a relationship between sleep quality and different occupations?” This is relevant because some jobs are may be more demanding or stressful, potentially impacting sleep patterns. Previous studies have shown mixed results, examining short sleep prevalence, working hours, and occupational stress. We will use linear regression models and plots to detect trends and statistically significant relationships, aiming to uncover insights about how different careers may influence sleep.

## Data and Methodology
 In our research, we utilized a data set containing sleep and health/lifestyle data with 374 individual unique Person IDs. We have narrowed this information down to exclude occupations that we deem to contain too few entrants. We set the cutoff for this at 20, meaning, we removed occupations that made up around 5% or less of the population. This helped to ensure more accuracy in our models, and brought us down to 7 Occupations out of the original 10. This left us with a total of 363 rows of data, which we deemed sufficient for this study.Each remaining row contains multiple observations, collected at once from each individual. We specifically would like to focus on the Y-variable of sleep quality, as we felt we may see unexpected responses when compared to sleep duration that could create notable diversity between our occupations. Sleep quality is a 1-10 **Y** measurement that we will be treating as a continuous variable in this case. We will also be comparing captured sleep quality to our other X variables in our dataset as we explore.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#creation of exploration and confirmation set
invisible(capture.output({
  filtered_occ_data <- sleep_data %>%
  filter(!Occupation %in% c("Software Engineer", "Manager", "Scientist", "Sales Representative"))
  model_filtered <- multinom(Occupation ~ Sleep.Duration + Quality.of.Sleep + Stress.Level, 
                           data = filtered_occ_data)
}))

invisible(capture.output({
  filtered_occ_data$Predicted <- predict(model_filtered)

  model <- multinom(Occupation ~ Quality.of.Sleep, data = filtered_occ_data)


set.seed(123)

#Sample 30% from each occupation for exploration set
exploration_set <- filtered_occ_data %>%
  group_by(Occupation) %>%
  slice_sample(prop = 0.3) %>%
  ungroup()

#get the rest as the confirmation set
confirmation_set <- anti_join(filtered_occ_data, exploration_set, by = colnames(filtered_occ_data))

#Occupation is now a factor in both datasets
exploration_set$Occupation <- as.factor(exploration_set$Occupation)
confirmation_set$Occupation <- as.factor(confirmation_set$Occupation)

#Extract occupation levels from exploration_set
occupation_levels <- levels(exploration_set$Occupation)

#Create named contrast matrix
contrasts_matrix <- contr.sum(length(occupation_levels))

#Add row and column names
rownames(contrasts_matrix) <- occupation_levels
colnames(contrasts_matrix) <- occupation_levels[-length(occupation_levels)]  # drop last level (R does this by default)

#Apply named contrast matrix
contrasts(exploration_set$Occupation) <- contrasts_matrix

contrasts(confirmation_set$Occupation) <- contr.sum(n = length(levels(confirmation_set$Occupation)))

#Fit models
model_exploration <- multinom(Occupation ~ Sleep.Duration + Quality.of.Sleep + Stress.Level + Physical.Activity.Level, 
                              data = exploration_set)


model_exploration_sleep_qual_alone <- lm(Quality.of.Sleep ~ Occupation, data = exploration_set)


model_exploration_sleep_qual <- lm(Quality.of.Sleep ~ Sleep.Duration + Occupation + Stress.Level + Physical.Activity.Level, 
                              data = exploration_set)

model_exploration_sleep_duration_stress <- lm(Quality.of.Sleep ~ Sleep.Duration + Stress.Level, 
                              data = exploration_set)

#model_exploration_sleep_qual_alone <- multinom(Quality.of.Sleep ~ Occupation, 
                              #data = exploration_set)
#Occupation is not numeric
#model_exploration_occ <- lm(Occupation ~ Quality.of.Sleep, 
                             # data = exploration_set)



}))
```
  Here, we show a model giving a visual of our coefficients for quality of sleep based on occupation.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
summary(model_exploration_sleep_qual_alone)

```
  From this summary of our exploratory sleep quality model, we made some important tweaks. As we do not have a baseline occupation, we instead took a grand mean to use as our baseline, incorporating all occupations. This number comes to an average of 7.324 quality of sleep. Based off of this data alone, we see a significant p-value for Accountant(+0.58, p < 0.05), Doctor(-0.51, p < 0.01), and Lawyer(+0.60, p < 0.01), very strong significance in both Engineer(+1.12, p < 0.001), and Salesperson(–1.32, p < 0.001), and a value without statistical significance in Nurses. With their coefficients in mind, we can mixed, yet significant results. Our R squared gives us a 43% proportion of variance in sleep quality and an adjusted value of 39.5% explained by occupation by this slice of data and our number of predictors. With our F-statistic and p-value, we can convincingly imagine that in this data slice, Occupation is a significant predictor of quality of sleep.
```{r}

summary(model_exploration_sleep_qual)

```

```{r, message=FALSE, warning = FALSE, echo = FALSE}
#table(sleep_data$Occupation)
#table(sleep_data$Quality.of.Sleep)
#model_uno <- lm(Occupation ~ Quality.of.Sleep, data = filtered_occ_data)

#summary(model_uno)
#summary(model)
```
  With the addition of Stress Level, Physical Activity Level, and Sleep Duration, our model now explains 94.09% of sleep quality variance. We chose to omit age and gender here, and even the existence of sleep disorders. This slice shows that sleep duration (+0.40, p < 0.001) and stress level(–0.40, p < 0.001) contribute heavily to sleep quality,though physical activity does not appear significant here. There also seems to be a shift in the effect of occupation on sleep quality, with Doctor no longer appearing significant, and appearing to have a higher tendency to stress and/or have lower sleep duration independently. These could obviously be overlapping with occupation, but their addition enhances our model nonetheless.
```{r}
tidy_model <- tidy(model_exploration_sleep_qual, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    is_occupation = grepl("^Occupation", term)
  )

#Get reordered term levels
term_levels <- tidy_model %>% arrange(estimate) %>% pull(term)

#Create a color for Occ
label_colors <- ifelse(grepl("^Occupation ", term_levels), "red", "black")
names(label_colors) <- term_levels

fig1 <- ggplot(tidy_model, aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(aes(color = is_occupation)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high, color = is_occupation), width = 0.2) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"), guide = "none") +
  coord_flip() +
  labs(title = "Fig1: Effect of Predictors on Sleep Quality",
       x = "Predictor",
       y = "Coefficient Estimate") +
  theme_minimal() +
  theme(axis.text.y = element_text(color = label_colors))


fig2 <- ggplot(exploration_set, aes(x = Sleep.Duration, y = Quality.of.Sleep, color = Occupation)) +
  geom_jitter(width = 0.1, height = 0.1, alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Fig2: Sleep Quality/Occupation",
    x = "Sleep Duration (hours)",
    y = "Quality of Sleep"
  ) +
  theme_minimal()

fig1 / fig2 + 
  plot_layout(heights = c(1, 1))

```
Here is a visual(Figure 1 or Fig1) of our X-variables effects (coefficient) on our outcome variable. We see that sleep duration and stress level are most significant, but when directly evaluating for occupation (in red), Salesperson has a very significant negative correlation, and Lawyer as a very significant positive correlation.
We can see in Figure 2 (or Fig2), that there is in fact, a positive trend with sleep duration and sleep quality as expected, with some occupations clearly falling behind regardless. We use jittered points here to give a stronger representation of our data visually, and give a properly continuous look.

## Results
  With our confirmation set, we will test how well our exploration set can be used to predict our values.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
confirmation_set$Predicted <- predict(model_exploration, newdata = confirmation_set)

#Evaluate accuracy or confusion matrix
conf_matrix1 <- table(Predicted = confirmation_set$Predicted, Actual = confirmation_set$Occupation)


#Create exploration set of 30% from each Quality.of.Sleep group
exploration_set2 <- filtered_occ_data %>%
  group_by(Quality.of.Sleep) %>%
  slice_sample(prop = 0.3) %>%
  ungroup()


#Confirmation set
confirmation_set2 <- anti_join(filtered_occ_data, exploration_set2, by = colnames(filtered_occ_data))

all_levels <- as.character(1:10)
exploration_set2$Quality.of.Sleep <- factor(exploration_set2$Quality.of.Sleep, levels = all_levels)
confirmation_set2$Quality.of.Sleep <- factor(confirmation_set2$Quality.of.Sleep, levels = all_levels)


invisible(capture.output({
#Train
  model_exploration2 <- multinom(Quality.of.Sleep ~ Sleep.Duration + Occupation + Stress.Level + Physical.Activity.Level, 
                                data = exploration_set2)
  
    #model_exploration2_qual <- lm(Quality.of.Sleep ~ Occupation, 
                                #data = exploration_set2)
}))


#Predict
confirmation_set2$Predicted <- predict(model_exploration2, newdata = confirmation_set2)

confirmation_set2$Predicted <- factor(confirmation_set2$Predicted, levels = all_levels)

#confusion matrix rename
conf_matrix2 <- table(Predicted = confirmation_set2$Predicted, Actual = confirmation_set2$Quality.of.Sleep)
rownames(conf_matrix2) <- paste0(rownames(conf_matrix2), "_QoS")
colnames(conf_matrix2) <- paste0(colnames(conf_matrix2), "_QoS")

#conf_matrix1
conf_matrix2

```
  We can see here, that Quality of Sleep (QoS) also has strong predictability. This is a very good sign that our data is significant due to accurate predictions of sleep quality values using our exploration set.
```{r, echo=FALSE, message=FALSE, warning=FALSE}

invisible(capture.output({
  #Occupation not numeric
  summary(model_exploration2)
  #avPlots(model_exploration2)

  ## Correlation Matrix
  numeric_data <- exploration_set2 %>%
      transmute(
      Quality.of.Sleep = as.numeric(as.character(Quality.of.Sleep)),
      Sleep.Duration   = as.numeric(Sleep.Duration),
      Stress.Level     = as.numeric(Stress.Level),
      Physical.Activity.Level = as.numeric(Physical.Activity.Level)
  )

#cor_matrix <- cor(numeric_data)


  dat <- subset(filtered_occ_data, complete.cases(Quality.of.Sleep, Occupation))

  dat$Occupation <- factor(dat$Occupation)
  dat$Quality.of.Sleep <- as.numeric(as.character(dat$Quality.of.Sleep))
}))
# Fit ANOVA
fit <- aov(Quality.of.Sleep ~ Occupation, data = dat)

# F-test and p-value
#summary(fit)        
#pairwise comparisons between occupations
#TukeyHSD(fit)       

# Effect size (eta squared)
#ss <- summary(fit)[[1]][, "Sum Sq"]
#eta2 <- ss["Occupation"] / sum(ss)
#eta2

```

## Discussion
Our analysis found a meaningful relationship between occupation and self-reported sleep quality. While job title alone explained some variation, our stronger model, adjusting for sleep duration and stress offered a clearer picture. Sleep duration and stress level were the most influential predictors, but occupation remained an important factor.
Notably, Salespeople consistently reported significantly lower sleep quality across both models. This suggests that unmeasured factors such as irregular hours, travel, or performance pressure may uniquely impact sleep in this group. Future research could explore these dimensions more closely. Additionally, it is important to note that  the data is entirely self-reported. Differences in how individuals perceive and report sleep and stress may vary by occupation, introducing bias. Incorporating objective measures in future studies would help validate these findings.
In sum, while stress and sleep duration strongly influence sleep quality and likely have occupational overlap, occupation itself continues to matter, suggesting it’s a valuable lens for understanding sleep health in the workforce.


\newpage

## Appendix
https://www.kaggle.com/datasets/uom190346a/sleep-health-and-lifestyle-dataset

Github: https://github.com/kadinw/Project2-MIDS203/tree/dev

For a first model we chose Occupation and Sleep quality due to the fact that these are the two variables we are most interested in seeing the relationship between, and the core of the study.

For a second model we used sleep duration, sleep quality, occupation, and stress level because we believed that stress level and sleep duration could potentially be more effective in predicting sleep quality than occupation, making it a reliable way to test our hypothesis.

A third model was also created, showing high sleep quality predictive value on sleep duration and stress level, but due to the focus of our report being occupation, it was not included


```{r}
exploration_set$fitted <- predict(model_exploration_sleep_qual)
exploration_set$residuals <- residuals(model_exploration_sleep_qual)

fig3 <- ggplot(exploration_set, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.6, color = "green") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

fig3
```
\newpage

# References

Luckhaupt, S. E. (2010, February 1). OUP. Oxford Academic. https://academic.oup.com/sleep/article-abstract/33/1/29/2454494 

Kim, B. H., & Lee, H. E. (2015). The association between working hours and sleep disturbances according to occupation and gender. Chronobiology International, 32(8), 1109–1114. https://doi.org/10.3109/07420528.2015.1064440

Mao, Y., Raju, G., & Zabidi, M. A. (2023, November 14). Association between occupational stress and sleep quality: A systematic review. Nature and science of sleep. https://pmc.ncbi.nlm.nih.gov/articles/PMC10656850/ 

::: {#refs}
:::
